\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-08-21 ResponseToReview.tex

One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.

\res{We have added the Balsas River Basin in a revised version of Figure 1.} \jri{LV please add, don't wait for Matt.} \mbh{see the R tutorial I sent, LV}

\section*{Response to Reviewer 2} 
Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some teosinte individuals; however, previous studies sampled a small fraction of the populations and individuals per population that we present here and thus did not provide comprehensive estimates. 
Our study is not only concerned with the presence or absence of the \emph{Hopscotch} allele in these populations but also the frequency at which it segregates.  
This finer level of detail sheds light on the evolutionary history of the \emph{Hopscotch} insertion in teosinte. 
We have added clarifying text to the introduction: ``The effects of the \emph{Hopscotch} insertion have been studied in maize (Studer \emph{et al.} 2011), and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes (Studer and Doebley 2012), but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and landrace maize; yet little is known about the frequency at which the \emph{Hopscotch} is segregating in natural populations."}

### 2014-08-21 manuscript.tex

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 1110 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected from a single teosinte population where \emph{Hopscotch} was found segregating.

@@

\emph{Key results:} Genotyping results suggest the \emph{Hopscotch} element is at higher than expected frequency in teosinte. Analysis of linkage disequilibrium near \emph{tb1} does not support recent introgression of the \emph{Hopscotch} allele from maize into teosinte.  Population genetic signatures are consistent with selection on this locus revealing a potential ecological role for \emph{Hopscotch} in teosinte. Finally, a greenhouse experiment with teosinte does not suggest \emph{tb1} controls tillering in natural populations.
\vspace*{0.2cm}

@@

In teosinte and other plants that grow at high population density, individuals detect competition from neighbors via the ratio of red to far-red light.  An increase in far-red relative to red light accompanies shading and triggers the shade avoidance syndrome: a suite of physiological and morphological changes such as reduced tillering, increased plant height and early flowering \citep{KebromBrutnell2007}. The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} \citep{LukensDoebley1999} and other grasses \citep{KebromBrutnell2007} and may therefore be crucial to the ecology of teosinte. In this study we aim to characterize the distribution of the \emph{Hopscotch} insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the \emph{Hopscotch} element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the \emph{Hopscotch} element and tillering phenotypes in a population of \emph{parviglumis}.

@@

\textbf{Sampling and genotyping---}We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations; ranging from 1-38 individuals per population) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion. Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes are three-banded (Appendix 2, See Supplemental Materials with the online version of this article). If only one of the two PCR reactions resolved well, we scored one allele for the individual. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

>>>>

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 1110 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected using seed from a single teosinte population where \emph{Hopscotch} was found segregating.

@@

\emph{Key results:} Genotyping results suggest the \emph{Hopscotch} element is at higher than expected frequency in teosinte. Analysis of linkage disequilibrium near \emph{tb1} does not support recent introgression of the \emph{Hopscotch} allele from maize into teosinte.  Population genetic signatures are consistent with selection on this locus revealing a potential ecological role for \emph{Hopscotch} in teosinte. Finally, a greenhouse experiment with teosinte does not suggest \emph{tb1} controls tillering in teosinte.
\vspace*{0.2cm}

@@

In teosinte and other plants that grow at high population density, individuals detect competition from neighbors via the ratio of red to far-red light.  An increase in far-red relative to red light accompanies shading and triggers the shade avoidance syndrome: a suite of physiological and morphological changes such as reduced tillering, increased plant height and early flowering \citep{KebromBrutnell2007}. The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} \citep{LukensDoebley1999} and other grasses \citep{KebromBrutnell2007} and may therefore be crucial to the ecology of teosinte. In this study we aim to characterize the distribution of the \emph{Hopscotch} insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the \emph{Hopscotch} element in our full panel and sequencing of two small regions upstream of \emph{tb1} combined with a larger SNP dataset in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the \emph{Hopscotch} element and tillering phenotypes in a population of \emph{parviglumis}.

@@

\textbf{Sampling and genotyping---}We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations; ranging from 1-38 individuals per population) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion due to its large size ($\sim$5kb). Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). If only one of the two PCR reactions resolved well, we scored one allele for the individual. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.
